## Practical 2: Mixed Models for Continuous Data
We start by loading the packages we will need for this practical and the data from 
[GitHub](https://github.com/drizopoulos/Repeated_Measurements). This is achieved with the
following commands:
```{r}
# packages
library("lattice")
library("nlme")
library("splines")

# data
con <- url("https://raw.github.com/drizopoulos/Repeated_Measurements/master/Data.RData")
load(con)
close(con)
```
```{r, echo = FALSE}
options(warn = (-1))
```

### Question 1
We produce the plot of the subject-specific trajectories separately per treatment group with
a superimposed loess curve. This achieved with the following call to function 
[xyplot()](https://goo.gl/cGSjZk):
```{r}
length.no.NA <- function (x) sum(!is.na(x))
ns <- with(pbc2, tapply(prothrombin, id, length.no.NA))
ns <- names(ns)[ns >= 6]
```

```{r}
set.seed(123)
ids <- sample(ns, 16) 
xyplot(prothrombin ~ year | id,
       panel = function (x, y, ...) {
           panel.xyplot(x, y, type = "p", col = 1, ...)
           panel.loess(x, y, col = 2, lwd = 2, span = 0.8)
       }, data = pbc2, subset = id %in% ids, layout = c(4, 4), as.table = TRUE, 
       xlab = "Time (years)", ylab = "Prothrombin Time (sec)")
```

```{r}
set.seed(321)
ids <- sample(ns, 16) 
xyplot(prothrombin ~ year | id,
       panel = function (x, y, ...) {
           panel.xyplot(x, y, type = "p", col = 1, ...)
           panel.loess(x, y, col = 2, lwd = 2, span = 0.8)
       }, data = pbc2, subset = id %in% ids, layout = c(4, 4), as.table = TRUE, 
       xlab = "Time (years)", ylab = "Prothrombin Time (sec)")
```

Some observations

* There is considerable variability in the profiles of prothrombin time.
* There is an overall increasing trend indicating worsening of the condition of the 
patients.
* No apparent differences can be seen between the two treatment groups, and between sexes.


### Question 2

### Question 3

### Question 4

### Question 5

### Question 6

### Question 7




