## Practical 2: Mixed Effects Models for Continuous Data
We start by loading the packages we will need for this practical and the data from 
[GitHub](https://github.com/drizopoulos/Repeated_Measurements). This is achieved with the
following commands:
```{r}
# packages
library("lattice")
library("nlme")
library("splines")

# data
con <- url("https://raw.github.com/drizopoulos/Repeated_Measurements/master/Data.RData")
load(con)
close(con)
```
```{r, echo = FALSE}
options(warn = (-1))
```

### Question 1
The following piece of code extracts the number of repeated measurements per patient:
```{r}
length.no.NA <- function (x) sum(!is.na(x))
ns <- with(pbc2, tapply(prothrombin, id, length.no.NA))
```

And we produce numerical summaries and a visualization using a boxplot:
```{r}
c(summary(ns), sd = sd(ns))
boxplot(ns, col = "lightgrey", ylab = "Number Repeated Measurements per Patient")
```

The median number of repeated measurements per individual is five, and only 25% of the 
patients have less than three measurements. Hence, we have sufficient information to model
potential nonlinearities in the subject-specific trajectories.

### Question 2
We produce the plot of the subject-specific trajectories separately per treatment group 
with a superimposed loess curve. This plot is produced for patients who have more than six
measurements. The plot is then produced with the following call to 
function [xyplot()](https://goo.gl/cGSjZk) (the same code is run twice with different seeds
in order to get different samples of patients):
```{r}
set.seed(123)
ids <- sample(names(ns)[ns >= 6], 16) 
xyplot(prothrombin ~ year | id,
       panel = function (x, y, ...) {
           panel.xyplot(x, y, type = "p", col = 1, ...)
           panel.loess(x, y, col = 2, lwd = 2, span = 0.8)
       }, data = pbc2, subset = id %in% ids, layout = c(4, 4), as.table = TRUE, 
       xlab = "Time (years)", ylab = "Prothrombin Time (sec)")
```

```{r}
set.seed(321)
ids <- sample(names(ns)[ns >= 6], 16) 
xyplot(prothrombin ~ year | id,
       panel = function (x, y, ...) {
           panel.xyplot(x, y, type = "p", col = 1, ...)
           panel.loess(x, y, col = 2, lwd = 2, span = 0.8)
       }, data = pbc2, subset = id %in% ids, layout = c(4, 4), as.table = TRUE, 
       xlab = "Time (years)", ylab = "Prothrombin Time (sec)")
```
```{r, echo = FALSE}
rm(list = ".Random.seed", envir = globalenv())
```

### Question 3
To fit the linear mixed effects model we use function [lme()](https://goo.gl/4fngXl). The
basic call to this function requires to specify argument `fixed` which is a formula 
providing the fixed-effects structure, argument `random` which is a formula providing 
the random-effects structure, and `data` the data frame that contains the variables we 
need. Hence, the call to [lme()](https://goo.gl/4fngXl) to fit the mixed for prothrombin 
time with 

* fixed effects: linear and quadratic time evolutions, the nonlinear effect of age 
using natural cubic splines with 3 degrees of freedom, the main effects of sex and drug, 
and the interactions of time with sex and drug, and age with sex and drug, and

* random effects: random intercepts

is:

```{r}
fm_1 <- lme(prothrombin ~ (year + I(year^2)) * (sex + drug) + ns(age, 3) + 
                ns(age, 3) : (sex + drug), data = pbc2, 
            random = ~ 1 | id, subset = prothrombin < 18)
```

Note that we use the `subset` argument to exclude the prothrombin times which are larger 
than 18 sec.

### Question 4
We will start expanding the random-effects structure, while keeping the same fixed effects,
in order to appropriately model the correlations in the repeated prothrombin time 
measurements. We start by including the linear random slopes term; note that instead of
calling [lme()](https://goo.gl/4fngXl) again and having to specify all arguments, we just
use function `update()`:
```{r}
fm_2 <- update(fm_1, random = ~ year | id)
```

To test whether the random slopes terms improves the fit of the model we perform the 
likelihood ratio test using the [anova()](https://goo.gl/88h4GZ) function:
```{r}
anova(fm_1, fm_2)
```

The results suggest that this term significantly improves the fit. We continue by 
including the quadratic random slopes term, i.e.,
```{r, cache = TRUE}
fm_3 <- update(fm_1, random = ~ year + I(year^2) | id)
```

Again we test using the likelihood ratio test:
```{r}
anova(fm_2, fm_3)
```

The fit improves again significantly. We continue by adding the cubic random slopes term,
i.e.,
```{r, cache = TRUE}
fm_4 <- update(fm_1, random = ~ year + I(year^2) + I(year^3) | id)
```

And we test again using likelihood ratio test:
```{r}
anova(fm_3, fm_4)
```

From this last test we obtain virtually the same likelihood value, and a p-value almost 
equal to 1. Hence, the cubic random slopes term does not seem to further improve the fit, 
and therefore we will continue with model `fm_3`.

### Question 5
Having chosen the random-effects structure we turn now in the fixed effects. We will first
test whether the interaction terms have an important contribution in the fit of the model.
Due to the fact that we have also interaction of the time variable with baseline 
covariates (i.e., sex and drug) we cannot compute the F-test because the denominator 
degrees of freedom cannot be (reliably) defined. To this end, we will perform a 
likelihood ratio test. Remember, as we have seen in **Practical 1**, to perform a 
likelihood ratio test we need to fit the model under maximum likelihood (and not the 
restricted maximum likelihood-the default in [lme()](https://goo.gl/4fngXl)). Thus,
in the following piece of code we first re-fit model `fm_3` using maximum likelihood and
then we fit the model that excludes the interaction terms (note that we efficiently use
function `update()` again):
```{r, cache = TRUE}
fm_3ML <- update(fm_3, method = "ML")
fm_3ML_noInt <- update(fm_3ML, fixed = . ~ year + I(year^2) + sex + drug + ns(age, 3))
```

Then we can compare the two models using the [anova()](https://goo.gl/88h4GZ) function:
```{r}
anova(fm_3ML_noInt, fm_3ML)
```

The results suggest that the effects of time and age on the prothrombin time do not 
differ between males and females, and between treated and non-treated patients.

### Question 6
We continue by testing the nonlinear terms. Similarly to what we did for the interaction
terms, we first perform the omnibus test for all the nonlinear terms in the model. As a 
initial step we need to fit the model that excludes all nonlinear terms, i.e.,
```{r, cache = TRUE}
fm_3ML_noInt_AllLin <- update(fm_3ML, fixed = . ~ year + sex + drug + age)
```

Then we test using [anova()](https://goo.gl/88h4GZ) function:
```{r}
anova(fm_3ML_noInt_AllLin, fm_3ML_noInt)
```

We have found a significant difference between the two models, meaning that some of the 
nonlinear terms have an important contribution to the model. We will proceed to find which
nonlinear terms we need. First, we are going to test the quadratic term for the time 
effect. To do this, we fit the model under the corresponding null hypothesis (i.e., we
drop the quadratic term for time but we keep the nonlinear terms for age):
```{r, cache = TRUE}
fm_3ML_noInt_LinTime <- update(fm_3ML, fixed = . ~ year + sex + drug + ns(age, 3))
```

The corresponding likelihood ratio test gives:
```{r}
anova(fm_3ML_noInt_LinTime, fm_3ML_noInt)
```

The result is significant implying the quadratic term is required. We continue by testing
the nonlinear age effect, by fitting the model under the corresponding null hypothesis 
(i.e., we now keep the quadratic term for time and we drop the nonlinear terms for age):
```{r, cache = TRUE}
fm_3ML_noInt_LinAge <- update(fm_3ML, fixed = . ~ year + I(year^2) + sex + drug + age)
```

The corresponding likelihood ratio test gives:
```{r}
anova(fm_3ML_noInt_LinAge, fm_3ML_noInt)
```

We have not found a significant difference between the two models, which means that we 
did not find any support from the data for a nonlinear age effect. Hence, we are going to
continue using model `fm_3ML_noInt_LinAge`.

### Question 7
Using the [summary()](https://goo.gl/u69ACN) function we obtain a detailed output of our 
final model `fm_3ML_noInt_LinAge`:
```{r}
summary(fm_3ML_noInt_LinAge)
```

The interpretation of the regression coefficients is as follows:

* For patients of the same sex and age at baseline who receive the same drug, each
year the average prothrombin time increases by **NA** sec.

The estimated random effects covariance matrix and the induced correlation matrix are:
```{r}
getVarCov(fm_3ML_noInt_LinAge, individuals = 271)
```
```{r}
cov2cor(getVarCov(fm_3ML_noInt_LinAge, individuals = 271))
```

The estimated marginal covariance matrix and the induced correlation matrix are:
```{r}
getVarCov(fm_3ML_noInt_LinAge, individuals = 271, type = "marginal")
```
```{r}
cov2cor(getVarCov(fm_3ML_noInt_LinAge, individuals = 271, type = "marginal")[[1]])
```

It is useful to compare these matrices and the results we obtained from the mixed model
analysis to the results we have obtained from our final model in **Practical 1**. As you
will see the values AIC and BIC of the mixed models are several orders of the magnitude 
lower than the corresponding marginal model.


### Question 8
To compare the marginal and subject-specific predictions from model `fm_3ML_noInt_LinAge`
we will use function [fitted()](https://goo.gl/cCuO4T). Argument `level` of this function
specifies the type of fitted values we obtain. The relevant calls are (first we define
the version of the dataset the excludes the outliers):
```{r}
pbc2n <- pbc2[pbc2$prothrombin < 18, ]
pbc2n$fitted_marg <- fitted(fm_3ML_noInt_LinAge, level = 0) # marginal
pbc2n$fitted_subj <- fitted(fm_3ML_noInt_LinAge, level = 1) # subject-specific
```

Then we create the plot using a similar call to function [xyplot()](https://goo.gl/KhfGGL)
as in Section 3.4:
```{r}
ids <- c(133, 36, 180, 11, 168, 116, 70, 58, 82, 104, 
         43, 21, 101, 210, 176, 157)
xyplot(prothrombin + fitted_marg + fitted_subj ~ year | id, data = pbc2n,
       panel = function (x, y, ...) {
           x.mat <- matrix(x, ncol = 3)
           y.mat <- matrix(y, ncol = 3)
           panel.xyplot(x.mat[, 1], y.mat[, 1], type = "p", col = "black")
           panel.xyplot(x.mat[, 2], y.mat[, 2], type = "l", lwd = 2, col = "red")
           panel.xyplot(x.mat[, 3], y.mat[, 3], type = "l", lwd = 2, col = "blue")
       }, subset = id %in% ids, layout = c(4, 4), as.table = TRUE, 
       xlab = "Time (years)", ylab = "Prothrombin Time (sec)")
```

As we have seen in Section 3.4, also here we observe that the subject-specific predictions
are much closer to the actual data of each subject than the corresponding marginal 
predictions.


### Question 9
To create the effect plot we first need to define the function `effectPlotData()` which we
have introduced in Section 3.2. This function takes a fitted mixed model, the data frame 
`newdata` based on which the plot will be made, and the original data we used to fit the 
model, and returns the predictions for each row of `newdata` with the corresponding lower 
and upper limits of the 95% confidence intervals.
```{r}
effectPlotData <- function (object, newdata, orig_data) {
    form <- formula(object)
    namesVars <- all.vars(form)
    respVar <- namesVars[1]
    newdata[[respVar]] <- 0.01
    betas <- if (!inherits(object, "lme")) coef(object) else fixef(object)
    V <- if (inherits(object, "geeglm")) object$geese$vbeta else vcov(object)
    orig_data <- orig_data[complete.cases(orig_data[namesVars]), ]
    mfX <- model.frame(terms(form), data = orig_data)
    X <- model.matrix(attr(mfX, "terms"), newdata)
    pred <- c(X %*% betas)
    ses <- sqrt(diag(X %*% V %*% t(X)))
    newdata$pred <- pred
    newdata$low <- pred - 1.96 * ses
    newdata$upp <- pred + 1.96 * ses
    newdata
}
```

Next we define the data frame with the combination of values of the covariates for which
we would like to create the plot. In our example, we set a regular sequence of 25 values 
from 0 to 12 for the time variable, the levels of the sex variable, and the levels of the 
drug variable:
```{r}
newDF <- with(pbc2n, expand.grid(year = seq(0, 12, length.out = 25),
                                sex = levels(sex), drug = levels(drug)))
```

As requested in this Question, we need to set the age variable in the corresponding 
median of the four groups (i.e., males in placebo, females in placebo, males in active 
treatment & females in active treatment) from the original dataset. This is achieved with
the following piece of code:
```{r}
pbc2n.id <- pbc2n[!duplicated(pbc2n$id), ]
median_age <- with(pbc2n.id, tapply(age, list(sex, drug), median))
ns <- with(newDF, tapply(year, list(sex, drug), length))
newDF$age <- rep(c(median_age), c(ns))
```

The first line defines the version of the dataset that contains a single row per patient
(if we would calculate the median age in each group in the original dataset we would not
obtain a correct result because patients have different numbers of repeated measurements).
The second line computes the median age in each of the four groups. The third line 
computes the number of rows that correspond to each of the four groups in the data frame
`newDF`. The final line puts the median ages in the corresponding rows. The effect plot is 
created with the following call to [xyplot()](https://goo.gl/KhfGGL):
```{r}
xyplot(pred + low + upp ~ year | sex * drug,
       data = effectPlotData(fm_3ML_noInt_LinAge, newDF, pbc2n), 
       lty = c(1, 2, 2), col = c(2, 1, 1), lwd = 2, type = "l",
       xlab = "Follow-up time (years)",
       ylab = "Prothrombin Time (sec)")
```

### Question 10
To investigate whether the assumption for the residuals terms are satisfied we can inspect
the scatterplots of the residuals versus the fitted values. As we have seen in Section 
3.11, there are two types of residuals we can compute in mixed models, namely the 
*marginal residuals* and the *conditional residuals*. We start with the standardized 
conditional residuals and we produce the overall scatteplot of these residuals versus the 
fitted values, and then we split per sex and treatment allocation:
```{r}
plot(fm_3ML_noInt_LinAge, resid(., type = "p") ~ fitted(.), 
     type = c("p", "smooth"), lwd = 3)
```
```{r}
plot(fm_3ML_noInt_LinAge, resid(., type = "p") ~ fitted(.) | sex, 
     type = c("p", "smooth"), lwd = 3)
```
```{r}
plot(fm_3ML_noInt_LinAge, resid(., type = "p") ~ fitted(.) | drug, 
     type = c("p", "smooth"), lwd = 3)
```

We continue by producing the same scatterplots but for the marginal residuals now:
```{r}
plot(fm_3ML_noInt_LinAge, resid(., type = "p", level = 0) ~ fitted(., level = 0), 
     type = c("p", "smooth"), lwd = 3)
```
```{r}
plot(fm_3ML_noInt_LinAge, resid(., type = "p", level = 0) ~ fitted(., level = 0) | sex, 
     type = c("p", "smooth"), lwd = 3)
```
```{r}
plot(fm_3ML_noInt_LinAge, resid(., type = "p", level = 0) ~ fitted(., level = 0) | drug, 
     type = c("p", "smooth"), lwd = 3)
```

Some observations:

* the conditional residuals show small faint signs of (remaining) heteroscedasticity, with
more spread for larger fitted values. Though this behavior may be influenced by the 
couple outlying measurements that still remain in the dataset.

* the marginal residuals show a clear systematic trend, with (almost) only negative 
residuals for large fitted values. Even though this would seem as an alarming sign of 
model misfit, we should take it with a grain of salt. As we will see in Chapter 6, 
residual plots are susceptible to produce misleading conclusions due to incomplete data
(i.e., dropout).
