## Practical 4: Mixed Effects Models for Discrete Data
We start by loading the packages we will need for this practical and the data from 
[GitHub](https://github.com/drizopoulos/Repeated_Measurements). This is achieved with the
following commands:
```{r, message = FALSE}
# packages
library("lattice")
library("lme4")
library("MASS")
library("splines")

# data
con <- url("https://raw.github.com/drizopoulos/Repeated_Measurements/master/Data.RData")
load(con)
close(con)
```
```{r, echo = FALSE}
options(warn = (-1))
```

### Question 1
The following piece of code creates the dummy variable for abnormal prothrombin time:
```{r}
pbc2$Dicht_prothro <- as.numeric(with(pbc2, prothrombin < 11 | prothrombin > 13))
```

### Question 2
To examine graphically the shape of the probability of abnormal prothrombin time we can
smooth the scatterplot of the `Dicht_prothro` for each patient separately. To obtain more
useful information we do that only in patients with five or more measurements. The 
following line of code creates an indicator variable denoting the rows of the `pbc2` 
dataset that correspond to patient with five measurements or more:
```{r}
ind <- with(pbc2, ave(Dicht_prothro, id, FUN = length)) >= 5
```

Next, using function [xyplot()](https://goo.gl/cGSjZk), we plot the smoothed 
subject-specific profiles of abnormal prothrombin time -- here we have opted for option 
`"spline"` in the `type` argument of [xyplot()](https://goo.gl/cGSjZk):
```{r}
xyplot(Dicht_prothro ~ year | id, data = pbc2, subset = ind,
       type = c("p", "spline"), lwd = 2,
       xlab = "Time (years)", ylab = "Abnormal Prothrombin Time",
       layout = c(6, 6), as.table = TRUE,
       scales = list(y = list(at = 0:1, labels = c("No", "Yes"))))
```

***

### Question 3
Based on the description of Question 3, we need to fit a mixed-effects logistic regression
model with random intercepts, and include in the mean structure the linear effect
of time, the main effect of sex and drug, the interaction of time with sex and the 
interaction effect of drug with sex. This model is fitted with following call to function
[glmer()](https://goo.gl/U7tf8T) (**note:** we also need to set the `family` argument to
`binomial()`, and the `nAGQ` argument to 15 to approximate the integrals with 15 
quadrature points):
```{r}
fm_1 <- glmer(Dicht_prothro ~ year * sex + drug + drug:sex + (1 | id), 
              data = pbc2, family = binomial(), nAGQ = 15)

summary(fm_1)
```

### Question 4
We continue the analysis by re-fitting the model of Question 3 by also including a random 
slopes term. This is done with the following call to [glmer()](https://goo.gl/U7tf8T) 
(**note:** the current version of [glmer()](https://goo.gl/U7tf8T) only allows for the 
adaptive Gaussian quadrature rule if only a single random effect is included; for more 
than one only the Laplace approximation is available):
```{r}
fm_2 <- glmer(Dicht_prothro ~ year * sex + drug + drug:sex + (year | id), 
              data = pbc2, family = binomial(), nAGQ = 1)
```

The likelihood ratio test is performed using the `anova()`:
```{r}
anova(fm_1, fm_2)
```

The results indicate that the random slopes is required. Nonetheless, due to the fact 
mixed models for categorical grouped data are more susceptible to numerical problems, it
is in general a good idea not only to look at the p-value from the likelihood ratio test,
but also at the estimated variance components. By doing this you could pick-up pathological
situations in which some variance parameters have converged on the boundary (i.e., very 
close to zero). Hence, we also inspect the `summary()` of the fitted model with random
intercepts and random slopes:
```{r}
summary(fm_2)
```

The estimated variances for the random intercepts and random slopes terms do not indicate
convergence on the boundary.

### Question 5
Next, we attempt to simplify the model by dropping all interaction terms. The additive 
mixed effects logistic regression model with only main effects is fitted using the 
following call to [glmer()](https://goo.gl/U7tf8T):
```{r}
fm_3 <- glmer(Dicht_prothro ~ year + sex + drug + (year | id), 
              data = pbc2, family = binomial(), nAGQ = 1)
```

Again the likelihood ratio test is calculated using the `anova()` function:
```{r}
anova(fm_3, fm_2)
```

The resulting p-value suggests that we can the simpler additive model.

### Question 6
The estimated fixed-effects coefficients from our final model are:
```{r}
summary(fm_3)
```


### Question 7

```{r}
effectPlotData_lmer <- function (object, newdata, orig_data, 
                                 type = c("lp", "response"), M = 100) {
    type <- match.arg(type)
    form <- formula(object)
    namesVars <- all.vars(form)
    fam <- family(object)
    orig_data <- orig_data[complete.cases(orig_data[namesVars]), ]
    TermsX <- delete.response(terms(object, fixed.only = TRUE))
    mfX <- model.frame(TermsX, data = orig_data)
    TermsX_new <- attr(mfX, "terms")
    betas <- fixef(object)
    V <- vcov(object)
    if (type == "lp" || (fam$family == "gaussian" && fam$link == "indentity")) {
        mfX_new <- model.frame(TermsX_new, newdata, xlev = .getXlevels(TermsX, mfX))
        X <- model.matrix(TermsX_new, mfX_new)
        eta <- c(X %*% betas)
        ses <- sqrt(diag(X %*% V %*% t(X)))
        newdata$pred <- eta
        newdata$low <- eta - 1.96 * ses
        newdata$upp <- eta + 1.96 * ses
        newdata
    } else {
        idVar <- names(object@flist)
        if (length(idVar) > 1)
            stop("The current version of this function only works for ",
                 "a single grouping factor.")
        ind <- rep(1:nrow(newdata), each = M)
        newdata2 <- newdata[ind, ]
        newdata2[[idVar]] <- factor(1:nrow(newdata2))
        mfX_new <- model.frame(TermsX_new, newdata2, xlev = .getXlevels(TermsX, mfX))
        X <- model.matrix(TermsX_new, mfX_new)
        formRE <- as.character(formula(object, random.only = TRUE))
        formRE <- as.formula(paste(formRE[c(1, 3)]))
        newRE <- lme4:::mkNewReTrms(object, newdata2, re.form = formRE,
                                    allow.new.levels = TRUE)
        D <- VarCorr(object)[[1]]
        b <- c(t(mvrnorm(nrow(newdata2), rep(0, NROW(D)), D)))
        eta0 <- c(X %*% betas)
        eta <- c(X %*% betas) + colSums(newRE$Zt * b)
        newdata$pred0 <- tapply(fam$linkinv(eta0), ind, mean)
        newdata$pred <- tapply(fam$linkinv(eta), ind, mean)
        newdata
    }
}
```

```{r}
newDF <- with(pbc2, expand.grid(
    year = seq(0, 12, length.out = 15),
    drug = levels(drug),
    sex = levels(sex)
))
```

```{r}
xyplot(pred + low + upp ~ year | sex * drug, 
       data = effectPlotData_lmer(fm_3, newDF, orig_data = pbc2), 
       type = "l", lty = c(1, 2, 2), col = c(2, 1, 1), lwd = 2)
```


### Question 8

```{r}
key <- simpleKey(c("marginal probabilities", "median patient"), points = FALSE, lines = TRUE)
key$lines$col <- c("red", "blue")
key$lines$lwd <- c(2, 2)
```

```{r}
xyplot(pred + pred0 ~ year | sex * drug, 
       data = effectPlotData_lmer(fm_3, newDF, orig_data = pbc2, 
                                  type = "response", M = 3000), 
       type = "l", lty = 1, lwd = 2, col = c("red", "blue"), key = key)
```
