# Practical 1: Marginal Models for Continuous Data
We start by loading the packages we will need for this practical and the data from 
[GitHub](https://github.com/drizopoulos/Repeated_Measurements). This is achieved with the
following commands:
```{r}
# packages
library("lattice")
library("nlme")
library("splines")

# data
con <- url("https://raw.github.com/drizopoulos/Repeated_Measurements/master/Data.RData")
load(con)
close(con)
```

## Question 1
We produce the plot of the subject-specific trajectories separately per treatment group with
a superimposed loess curve. This achieved with the following call to function 
[xyplot()](https://goo.gl/cGSjZk):
```{r}
xyplot(prothrombin ~ year | drug, group = id, data = pbc2, 
       panel = function (x, y, ...) {
           panel.xyplot(x, y, type = "l", col = 1, ...)
           panel.loess(x, y, col = 2, lwd = 2)
       }, xlab = "Time (years)", ylab = "Prothrombin Time (sec)")
```

With virtually an analogous call to [xyplot()](https://goo.gl/cGSjZk) we also produce the 
plot of the subject-specific trajectories per sex:
```{r}
xyplot(prothrombin ~ year | sex, group = id, data = pbc2, 
       panel = function (x, y, ...) {
           panel.xyplot(x, y, type = "l", col = 1, ...)
           panel.loess(x, y, col = 2, lwd = 2)
       }, xlab = "Time (years)", ylab = "Prothrombin Time (sec)")
```

From both plots we observe that we have some outlying measurements that affect the scale
of the y-axis. We could zoom in by specifying the `ylim` argument, e.g.,
```{r}
xyplot(prothrombin ~ year | drug, group = id, data = pbc2, 
       panel = function (x, y, ...) {
           panel.xyplot(x, y, type = "l", col = 1, ...)
           panel.loess(x, y, col = 2, lwd = 2)
       }, xlab = "Time (years)", ylab = "Prothrombin Time (sec)",
       ylim = c(9, 18))
```

```{r}
xyplot(prothrombin ~ year | sex, group = id, data = pbc2, 
       panel = function (x, y, ...) {
           panel.xyplot(x, y, type = "l", col = 1, ...)
           panel.loess(x, y, col = 2, lwd = 2)
       }, xlab = "Time (years)", ylab = "Prothrombin Time (sec)",
       ylim = c(9, 18))
```

Some observations

* There is considerable variability in the profiles of prothrombin time.
* There is an overall increasing trend indicating worsening of the condition of the 
patients.
* No apparent differences can be seen between the two treatment groups, and between sexes.


## Question 2
A basic all to function [gls()](https://goo.gl/0KBq4B) without specifying a correlation 
structure (i.e., argument `correlation`) or a variance function (i.e., argument `weights`)
fits a simple linear regression model (i.e., it is equivalent to 
[lm()](https://goo.gl/tewND3)). To include the natural cubic splines of time with 3 
degrees of freedom we need use function [ns()](https://goo.gl/0g115B). Hence, the call to
[gls()](https://goo.gl/0KBq4B) that includes the nonlinear time effect, the main effect 
of age, sex and drug, and the interaction terms between the nonlinear time and sex and 
drug is:
```{r}
fm_1 <- gls(prothrombin ~ ns(year, 3) * (sex + drug) + age, data = pbc2)

summary(fm_1)
```

Interpretation of the results

* The terms that involve the spline coefficients (main of time and its interaction with 
sex and drug) do not have a straightforward interpretation. This is where the effect plots
we have seen in Section 2.4 are of great use.
* The main effects of age, sex, and drug do have a meaningful interpretation. In 
particular:
    + The coefficient of age `r round(coef(fm_1)['age'], 4)` denotes the increase in 
  prothrombin time per year increase in age for patients of the same sex and who receive
  the same drug, and are compared at the same follow-up time.
    + The coefficient of drug `r round(coef(fm_1)['drugD-penicil'], 2)` denotes the difference
  in prothrombin time between the two treatment groups at baseline for patients of the same
  sex and age.

We should **not** simplify the model at this stage. As we have mentioned in Section 2.8, we
should first decide on the covariance structure while keeping the mean structure as 
general as possible.

## Question 3
The following call to `gls()` fit the
```{r}
fm_2 <- gls(prothrombin ~ ns(year, 3) * (sex + drug) + age, data = pbc2, 
            correlation = corCompSymm(form = ~ year | id))

fm_3 <- gls(prothrombin ~ ns(year, 3) * (sex + drug) + age, data = pbc2, 
            correlation = corCAR1(form = ~ year | id))

fm_4 <- gls(prothrombin ~ ns(year, 3) * (sex + drug) + age, data = pbc2, 
            correlation = corExp(form = ~ year | id))

fm_5 <- gls(prothrombin ~ ns(year, 3) * (sex + drug) + age, data = pbc2, 
            correlation = corGaus(form = ~ year | id))
```

```{r}
fm_6 <- gls(prothrombin ~ ns(year, 3) * (sex + drug) + age, data = pbc2, 
            correlation = corCompSymm(form = ~ year | id),
            weights = varExp(form = ~ year))

fm_7 <- gls(prothrombin ~ ns(year, 3) * (sex + drug) + age, data = pbc2, 
            correlation = corCAR1(form = ~ year | id),
            weights = varExp(form = ~ year))

fm_8 <- gls(prothrombin ~ ns(year, 3) * (sex + drug) + age, data = pbc2, 
            correlation = corExp(form = ~ year | id),
            weights = varExp(form = ~ year))

fm_9 <- gls(prothrombin ~ ns(year, 3) * (sex + drug) + age, data = pbc2, 
            correlation = corGaus(form = ~ year | id),
            weights = varExp(form = ~ year))
```

## Question 4

```{r}
anova(fm_1, fm_2)
anova(fm_1, fm_3)
anova(fm_1, fm_4)
anova(fm_1, fm_5)
```

```{r}
anova(fm_2, fm_3, fm_4, fm_5)
```

```{r}
anova(fm_6, fm_7, fm_8, fm_9)
```

```{r}
anova(fm_2, fm_6)
```

## Question 5

```{r}
anova(fm_6, Terms = c('ns(year, 3):sex', 'ns(year, 3):drug'))
```

```{r}
fm_6_noInt <- gls(prothrombin ~ ns(year, 3) + sex + drug + age, data = pbc2, 
                  correlation = corCompSymm(form = ~ year | id),
                  weights = varExp(form = ~ year))

anova(fm_6_noInt, fm_6)
```

```{r}
fm_6ML <- gls(prothrombin ~ ns(year, 3) * (sex + drug) + age, data = pbc2, 
            correlation = corCompSymm(form = ~ year | id),
            weights = varExp(form = ~ year), method = "ML")

fm_6ML_noInt <- gls(prothrombin ~ ns(year, 3) + sex + drug + age, data = pbc2, 
                  correlation = corCompSymm(form = ~ year | id),
                  weights = varExp(form = ~ year), method = "ML")

anova(fm_6ML_noInt, fm_6ML)
```

## Question 6

```{r}
fm_6ML_noInt <- gls(prothrombin ~ ns(year, 3) + sex + drug + age, data = pbc2, 
                  correlation = corCompSymm(form = ~ year | id),
                  weights = varExp(form = ~ year), method = "ML")

fm_6ML_noIntLin <- gls(prothrombin ~ year + sex + drug + age, data = pbc2, 
                       correlation = corCompSymm(form = ~ year | id),
                       weights = varExp(form = ~ year), method = "ML")

anova(fm_6ML_noIntLin, fm_6ML_noInt)
```

## Question 7

```{r}
summary(fm_6ML_noIntLin)
```

```{r}
getVarCov(fm_6ML_noIntLin, individual = 271)

cov2cor(getVarCov(fm_6ML_noIntLin, individual = 271))
```

## Question 8

```{r}
effectPlotData <- function (object, newdata, orig_data) {
    form <- formula(object)
    namesVars <- all.vars(form)
    respVar <- namesVars[1]
    newdata[[respVar]] <- 0.01
    betas <- if (!inherits(object, "lme")) coef(object) else fixef(object)
    V <- if (inherits(object, "geeglm")) object$geese$vbeta else vcov(object)
    orig_data <- orig_data[complete.cases(orig_data[namesVars]), ]
    mfX <- model.frame(terms(form), data = orig_data)
    X <- model.matrix(attr(mfX, "terms"), newdata)
    pred <- c(X %*% betas)
    ses <- sqrt(diag(X %*% V %*% t(X)))
    newdata$pred <- pred
    newdata$low <- pred - 1.96 * ses
    newdata$upp <- pred + 1.96 * ses
    newdata
}
```

```{r}
# the data frame that contains the combination of values to
# create the plot
newDF <- with(pbc2, expand.grid(year = seq(0, 12, length.out = 25),
                                sex = levels(sex), drug = levels(drug), age = 49))
```

```{r}
# the effects plot
xyplot(pred + low + upp ~ year | sex*drug, 
       data = effectPlotData(fm_6ML_noIntLin, newDF, pbc2), 
       lty = c(1, 2, 2), col = c(2, 1, 1), lwd = 2, type = "l",
       xlab = "Follow-up time (years)", ylab = "Prothrombin Time (sec)")
```