## Practical 3: Marginal Models for Discrete Data
We start by loading the packages we will need for this practical and the data from 
[GitHub](https://github.com/drizopoulos/Repeated_Measurements). This is achieved with the
following commands:
```{r}
# packages
library("lattice")
library("geepack")
library("splines")

# data
con <- url("https://raw.github.com/drizopoulos/Repeated_Measurements/master/Data.RData")
load(con)
close(con)
```
```{r, echo = FALSE}
options(warn = (-1))
```

### Question 1
The following piece of code creates the dummy variable for abnormal prothrombin time:
```{r}
pbc2$Dicht_prothro <- as.numeric(with(pbc2, prothrombin < 11 | prothrombin > 13))
```

### Question 2
To examine graphically the shape of the probability of abnormal prothrombin time we can
smooth the scatterplot of the `Dicht_prothro`. Using function 
[xyplot()](https://goo.gl/cGSjZk) as in previous practicals, we first create the 
scatterplot of `Dicht_prothro` versus follow-up time, separately for placebo and 
D-penicillamine patients, and we include the loess curve:
```{r}
xyplot(Dicht_prothro ~ year | drug, data = pbc2,
       type = c("p", "smooth"), lwd = 2,
       xlab = "Time (years)", ylab = "Abnormal Prothrombin Time",
       scales = list(y = list(at = 0:1, labels = c("No", "Yes"))))
```

Note that due to the binary nature of `Dicht_prothro`, the actual data points do not 
provide any useful information. However, the loess smoother can still be used and recover
the shape of the probability of abnormal prothrombin time (**note:** the probability of
abnormal prothrombin time is actually the sample mean of the `Dicht_prothro` variable). 
For both treatment groups we observe a decrease in the probability of abnormal prothrombin
time during follow-up. We continue by producing the same plot but now splitting the panels
for males and females:
```{r}
xyplot(Dicht_prothro ~ year | sex, data = pbc2,
       type = c("p", "smooth"), lwd = 2,
       xlab = "Time (years)", ylab = "Abnormal Prothrombin Time",
       scales = list(y = list(at = 0:1, labels = c("No", "Yes"))))
```

This figure suggests that males have more stable evolutions, and that actually the females
are the ones who improve over time. Note that at this stage this is just an observation 
from just looking at the data. To make a formal conclusion we will need to test this 
conjecture. We finally produce the same scatterplot and we split at different age 
categories, namely, [25, 43], [43, 50], [50, 55], and [55, 80]. To achieve this we use 
function [cut()](https://goo.gl/t8xP1d), i.e.,
```{r}
xyplot(Dicht_prothro ~ year | cut(age, c(25, 43, 50, 55, 80)), data = pbc2,
       type = c("p", "smooth"), lwd = 2, as.table = TRUE,
       xlab = "Time (years)", ylab = "Abnormal Prothrombin Time",
       scales = list(y = list(at = 0:1, labels = c("No", "Yes"))))
```

We observe some subtle difference in the evolutions of the different age categories.

### Question 3
Based on the description of Question 3, we need to fit a GEE model with an exchangeable
working correlation matrix, and include in the mean structure the nonlinear effect
of time using natural cubic splines with 2 degrees of freedom (i.e., using function 
[ns()](https://goo.gl/zeAjzF)), the interaction of this nonlinear effect with drug and 
sex (while also including the main effects of these variables), the main effect of age, 
and its interaction with sex. This model is fitted with following call to function
[geeglm()](https://goo.gl/1S1uix) (**note:** we also need to set the `family` argument to
`binomial()`, and supply the `id` argument to identify the clusters, i.e., in this case 
the patients):
```{r}
fm_1 <- geeglm(Dicht_prothro ~ ns(year, 2) * (drug + sex) + age + age:sex, 
               family = binomial(), data = pbc2, id = id, corstr = "exchangeable")

summary(fm_1)
```

The `summary()` function produces a detailed output of the fitted model, with the 
parameter estimates, standard errors and corresponding p-values.

### Question 4
We continue the analysis by re-fitting the model of Question 3 with the independence and 
auto-regressive of order 1 working correlation matrices. This is done with the following 
call to [geeglm()](https://goo.gl/1S1uix) by suitably changing the `corstr` argument:
```{r}
fm_2 <- geeglm(Dicht_prothro ~ ns(year, 2) * (drug + sex) + age + age:sex, 
               family = binomial(), data = pbc2, id = id, corstr = "independence")

fm_3 <- geeglm(Dicht_prothro ~ ns(year, 2) * (drug + sex) + age + age:sex, 
               family = binomial(), data = pbc2, id = id, corstr = "ar1")
```

We collect the estimated coefficients the two types of standard errors (i.e., sandwich and
naive) using similar code as in Section 4.5
```{r}
# parameters
betas <- cbind("independence" = coef(fm_2), 
            "exchangeable" = coef(fm_1),
            "AR1" = coef(fm_3))

# Sandwich standard errors
extractSEs <- function (model) sqrt(diag(model$geese$vbeta))
ses_sand <- cbind("independence" = extractSEs(fm_2), 
            "exchangeable" = extractSEs(fm_1),
            "AR1" = extractSEs(fm_3))

# Naive standard errors
extractSEs_naive <- function (model) sqrt(diag(model$geese$vbeta.naiv))
ses_naiv <- cbind("independence" = extractSEs_naive(fm_2), 
            "exchangeable" = extractSEs_naive(fm_1),
            "AR1" = extractSEs_naive(fm_3))
```

To make the coefficients' plot, we put the estimates and the calculated confidence 
intervals (both naive and robust) in a data frame:
```{r}
dat <- data.frame(
    est = c(betas), 
    lower_sand = unname(c(betas) - 1.96 * c(ses_sand)),
    lower_naiv = unname(c(betas) - 1.96 * c(ses_naiv)),
    upper_sand = unname(c(betas) + 1.96 * c(ses_sand)),
    upper_naiv = unname(c(betas) + 1.96 * c(ses_naiv)),
    parm = gl(11, 1, 33, labels = rownames(betas)),
    work_corr = gl(3, 11, labels = colnames(betas))
)
```

To produce the plot, we first define the `prepanel.ci2()` and `panel.ci2()` functions:
```{r}
prepanel.ci2 <- function (x, y, lx, ux, lx2, ux2, subscripts, ...) {
    x <- as.numeric(x)
    lx <- as.numeric(lx[subscripts])
    ux <- as.numeric(ux[subscripts])
    lx2 <- as.numeric(lx2[subscripts])
    ux2 <- as.numeric(ux2[subscripts])
    list(xlim = range(x, ux, lx, ux2, lx2, finite = TRUE))
}
panel.ci2 <- function (x, y, lx, ux, lx2, ux2, subscripts, pch = 16, ...) {
    x <- as.numeric(x)
    y <- as.numeric(y)
    lx <- as.numeric(lx[subscripts])
    ux <- as.numeric(ux[subscripts])
    lx2 <- as.numeric(lx2[subscripts])
    ux2 <- as.numeric(ux2[subscripts])
    panel.abline(h = c(unique(y)), 
                 col = "grey", lty = 2, lwd = 1.5)
    panel.arrows(lx, y, ux, y,
                 length = 0.1, unit = "native",
                 angle = 90, code = 3, lwd = 2, col = "blue")
    panel.arrows(lx2, y + 0.08, ux2, y + 0.08,
                 length = 0.1, unit = "native",
                 angle = 90, code = 3, lwd = 2, col = "magenta2")
    panel.xyplot(x, y, pch = pch, col = 2, cex = 0.9, ...)
    panel.xyplot(x, y + 0.08, pch = pch, col = 2, cex = 0.9, ...)
}
```

And following we call [dotplot()](https://goo.gl/7w7nrR):
```{r}
dotplot(work_corr ~  est | parm, lx = dat$lower_sand, lx2 = dat$lower_naiv, 
        ux = dat$upper_sand, ux2 = dat$upper_naiv,
        data = dat, xlab = "", prepanel = prepanel.ci2, panel = panel.ci2, 
        as.table = TRUE, scales = list(x = list(relation = "free")),
        key = simpleKey(c("Robust Standard Errors", "Naive Standard Errors"), 
                        points = FALSE, lines = TRUE, col = c("blue", "magenta2")))
```

A couple of observations:

* There are small differences in the point estimates of the corresponding parameters 
across the different models.

* For some parameters (e.g., `ns(year, 2)2:drugD-penicil` and `ns(year, 2)1:sexfemale`) 
we obtain a different conclusion regarding to their significance depending on the choice 
of the working correlation matrix.

* For some parameters, there are also noticeable differences between the naive and robust
standard errors.

<span style="color:red"><strong>Choice of the working correlation matrix:</strong></span>
As we have explained in the course, there is no (easy) formal way to test which working 
correlation matrix provides a better fit. Hence, the choice should be based on features of
the data we are studying. Given that PBC is a longitudinal study, it is reasonable to 
assume that abnormal prothrombin times that are closer in follow-up time will be more 
strongly correlated than abnormal prothrombin times that are further apart. With this in
mind, the AR1 working correlation matrix seems more appropriate and therefore we will 
continue with this matrix for the rest of this practical.

### Question 5
Based on the requests of the researchers from this study, we first attempt to simplify the
model by excluding interaction terms. We start with the omnibus test for all the 
interaction terms in the model. In the framework of GEE models we do not perform likelihood
ratio tests (because actually there is not likelihood function) but rather Wald tests. 
Even though the test is different, the implementation in R with the **geepack** package is
the same, i.e., we first need to fit the model under the corresponding null hypothesis, 
and the compare the model under the null (here the one the excludes all interaction terms)
with the model under the alternative (here the model that includes all interaction terms)
using the `anova()` function. The corresponding code is:
```{r}
fm_3_IntAll <- geeglm(Dicht_prothro ~ ns(year, 2) + drug + sex + age,
                      family = binomial(), data = pbc2, id = id, corstr = "ar1")

anova(fm_3_IntAll, fm_3)
```

We observe that the result is on the borderline. Nonetheless, we will continue to 
investigate the interaction terms further. We will first test the two interaction terms 
for the time effect, i.e.,
```{r}
fm_3_IntTimeAll <- geeglm(Dicht_prothro ~ ns(year, 2) + sex + drug + age + age : sex, 
                          family = binomial(), data = pbc2, id = id, corstr = "ar1")

anova(fm_3_IntTimeAll, fm_3)
```

We find a significant difference between the two models suggesting that at least one of 
the interaction terms is required. To find out which, we first test the interaction term 
for drug, i.e.,
```{r}
fm_3_IntTimeDrug <- geeglm(Dicht_prothro ~ ns(year, 2) * sex + drug + age + age : sex, 
                           family = binomial(), data = pbc2, id = id, corstr = "ar1")

anova(fm_3_IntTimeDrug, fm_3)
```

The results point out that this interaction terms does not seem to improve the fit. We 
continue with the interaction term for sex
```{r}
fm_3_IntTimeSex <- geeglm(Dicht_prothro ~ ns(year, 2) * drug + sex + age + age : sex, 
                          family = binomial(), data = pbc2, id = id, corstr = "ar1")

anova(fm_3_IntTimeSex, fm_3)
```

The interaction between the time effect and sex seems to improve the fit of the model. We
finally test the interaction term between age and sex
```{r}
fm_3_IntSexAge <- geeglm(Dicht_prothro ~ ns(year, 2) * (drug + sex) + age, 
                         family = binomial(), data = pbc2, id = id, corstr = "ar1")

anova(fm_3_IntSexAge, fm_3)
```

Hence, we keep as final model the one that only includes the interaction term between
sex and time, i.e.,
```{r}
fm_3_IntFinal <- geeglm(Dicht_prothro ~ ns(year, 2) * sex + drug + age, 
                        family = binomial(), data = pbc2, id = id, corstr = "ar1")
```


### Question 6
Next we would like to test the requirement to include the nonlinear terms. Using an
analogous procedure as in Question 5, we first fit the model under the null hypothesis, 
that is the model that assumes a linear effect for time, and we compare it with model
`fm_3_IntFinal` that assumes a nonlinear effect:
```{r}
fm_3_IntFinal_Lin <- geeglm(Dicht_prothro ~ year * sex + drug + age, 
                            family = binomial(), data = pbc2, id = id, corstr = "ar1")

anova(fm_3_IntFinal_Lin, fm_3_IntFinal)
```

The result suggests that we should retain the nonlinear effect.

### Question 7

```{r}
summary(fm_3_IntFinal)
```

Interpretation of the regression coefficients:

* The coefficients that involve the nonlinear terms of time do not have a straightforward
interpretation - in this cases an Effect Plot is more effective in communicating the 
results of the model (see next Question).

* The coefficients of drug and age can be interpreted in a easier manner, namely
    + `r round(coef(fm_3_IntFinal)['drugD-penicil'], 2)` is the log odds ratio between
    treated and placebo patients of the same age and sex, and who are compared at the same
    follow-up time. The corresponding odds ratio is 
    `r round(exp(coef(fm_3_IntFinal)['drugD-penicil']), 2)`, i.e., 
    $e^{`r round(coef(fm_3_IntFinal)['drugD-penicil'], 2)`}$.
    
    + `r round(coef(fm_3_IntFinal)['age'], 2)` is the log odds ratio for a unit increase
    in the baseline age for patients who have the same sex, received the same treatment,and
    who are compared at the same follow-up time. The corresponding odds ratio is 
    `r round(exp(coef(fm_3_IntFinal)['age']), 2)`, i.e., 
    $e^{`r round(coef(fm_3_IntFinal)['age'], 2)`}$.


### Question 8
To create the effect plot we first need to define the function `effectPlotData()` which we
have introduced in Section 4.3. This function takes a fitted GEE model, the data frame 
`newdata` based on which the plot will be made, and the original data we used to fit the 
model, and returns the predictions for each row of `newdata` with the corresponding lower 
and upper limits of the 95% confidence intervals.
```{r}
effectPlotData <- function (object, newdata, orig_data) {
    form <- formula(object)
    namesVars <- all.vars(form)
    betas <- if (!inherits(object, "lme")) coef(object) else fixef(object)
    V <- if (inherits(object, "geeglm")) object$geese$vbeta else vcov(object)
    orig_data <- orig_data[complete.cases(orig_data[namesVars]), ]
    Terms <- delete.response(terms(form))
    mfX <- model.frame(Terms, data = orig_data)
    Terms_new <- attr(mfX, "terms")
    mfX_new <- model.frame(Terms_new, newdata, xlev = .getXlevels(Terms, mfX))
    X <- model.matrix(Terms_new, mfX_new)
    pred <- c(X %*% betas)
    ses <- sqrt(diag(X %*% V %*% t(X)))
    newdata$pred <- pred
    newdata$low <- pred - 1.96 * ses
    newdata$upp <- pred + 1.96 * ses
    newdata
}
```

Next we define the data frame with the combination of values of the covariates for which
we would like to create the plot. In our example, we set a regular sequence of 25 values 
from 0 to 12 for the time variable, the levels of the sex variable, and the levels of the 
drug variable, and age at 49 years old:
```{r}
newDF <- with(pbc2, expand.grid(
    year = seq(0, 12, length.out = 25),
    age = 49,
    drug = levels(drug),
    sex = levels(sex)
))
```

The effect plot is created with the following call to [xyplot()](https://goo.gl/KhfGGL) -
first we produce the plot for the log odds of abnormal prothrombin time:
```{r}
xyplot(pred + low + upp ~ year | sex * drug, 
       data = effectPlotData(fm_3_IntFinal, newDF, pbc2), 
       lty = c(1, 2, 2), col = c(2, 1, 1), lwd = 2, type = "l",
       xlab = "Follow-up time (years)",
       ylab = "log Odds")
```

And following the plot for the probability of abnormal prothrombin time (the first line 
of code defines the helper function `expit()` which transform the log odds to 
probabilities):
```{r}
expit <- function (x) exp(x) / (1 + exp(x))
xyplot(expit(pred) + expit(low) + expit(upp) ~ year | sex * drug, 
       data = effectPlotData(fm_3_IntFinal, newDF, pbc2), 
       lty = c(1, 2, 2), col = c(2, 1, 1), lwd = 2, type = "l",
       xlab = "Follow-up time (years)",
       ylab = "Probabilities")
```

### Question 9
To test for differences for males and females patients at different time points we will
use function `rowDiff()` and the theory behind it, which we have introduced in Section 
4.6:
```{r}
rowDiff <- function (object, newdata, orig_data, adjust.p = FALSE, ...) {
    form <- formula(object)
    namesVars <- all.vars(form)
    betas <- if (!inherits(object, "lme")) coef(object) else fixef(object)
    V <- if (inherits(object, "geeglm")) object$geese$vbeta else vcov(object)
    orig_data <- orig_data[complete.cases(orig_data[namesVars]), ]
    Terms <- delete.response(terms(form))
    mfX <- model.frame(Terms, data = orig_data)
    Terms_new <- attr(mfX, "terms")
    mfX_new <- model.frame(Terms_new, newdata, xlev = .getXlevels(Terms, mfX))
    X <- model.matrix(Terms_new, mfX_new)
    ind <- combn(nrow(X), 2)
    k <- ncol(ind)
    out <- matrix(0, k, 5)
    for (i in seq_len(k)) {
        XX <- X[ind[1, i], , drop = FALSE] - X[ind[2, i], , drop = FALSE]
        est <- drop(XX %*% betas)
        se <- sqrt(diag(XX %*% V %*% t(XX)))
        out[i, 1] <- est
        out[i, 2] <- se
        out[i, 3] <- est - 1.96 * se
        out[i, 4] <- est + 1.96 * se
        out[i, 5] <- 2 * pnorm(abs(est / se), lower.tail = FALSE)
    }
    if (k > 2 && adjust.p) {
        out[, 5] <- p.adjust(out[, 5], ...)
    }
    colnames(out) <- c("Diff", "Std.Err.", "95%low", "95%upp", "p-value")
    rownames(out) <- paste(ind[1, ], "-", ind[2, ])
    out
}
```

To use this function we first need to define the data frame with rows the corresponding 
combinations of the covariates we would like to test. Hence, for the time variable we 
choose the follow-up years 2, 6, 8, and 10, for age 49 years old, from drug 
D-penicillamine, and for sex males:
```{r}
nDF_male <- with(pbc2, expand.grid(
    year = c(2, 6, 8, 10), age = 49,
    drug = levels(drug)[2],
    sex = levels(sex)[1]
))
nDF_male
```

The we the following call to `rowDiff()` we perform the pairwise comparisons:
```{r}
rowDiff(fm_3_IntFinal, nDF_male, pbc2)
```

Before going to interpret the results we should note that we have performed multiple tests
and therefore we should accordingly adjust the p-values. To do that we set argument
`adjust.p` of `rowDiff()` to `TRUE`, i.e.,
```{r}
rowDiff(fm_3_IntFinal, nDF_male, pbc2, adjust.p = TRUE)
```

Each row in the output of `rowDiff()` corresponds to a comparison of the corresponding rows 
of the data frame `nDF_male`, i.e., the first row is the comparison between year 2 and 
year 6, the second row between year 2 and year 8, and so on. After adjusting for multiple 
comparisons we do not find any significant results. We continue to perform the same
pairwise comparisons for the females. First, we construct the data frame `nDF_female`
accordingly:
```{r}
nDF_female <- with(pbc2, expand.grid(
    year = c(2, 6, 8, 10), age = 49,
    drug = levels(drug)[2],
    sex = levels(sex)[2]
))
nDF_female
```

And we use it in the call to `rowDiff()`:
```{r}
rowDiff(fm_3_IntFinal, nDF_female, pbc2, adjust.p = TRUE)
```

For females, and after adjusting for multiple comparisons we observe that all differences
are significant.