---
title: "Count Data Mixed Models - Extensions"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library("GLMMadaptive")
library("lattice")
```

```{r simulate_data, include = FALSE}
set.seed(1234)
n <- 300 # number of subjects
K <- 8 # number of measurements per subject
t_max <- 5 # maximum follow-up time

# we constuct a data frame with the design: 
# everyone has a baseline measurment, and then measurements at random follow-up times
DF <- data.frame(id = rep(seq_len(n), each = K),
                 time = c(replicate(n, c(0, sort(runif(K - 1, 0, t_max))))),
                 sex = rep(gl(2, n/2, labels = c("male", "female")), each = K),
                 age = rep(round(runif(n, 40, 75), 1), each = K))

# design matrices for the fixed and random effects non-zero part
X <- model.matrix(~ sex * time + age, data = DF)
Z <- model.matrix(~ time, data = DF)
# design matrices for the fixed and random effects zero part
X_zi <- model.matrix(~ sex + age, data = DF)
Z_zi <- model.matrix(~ 1, data = DF)

betas <- c(1.5, 0.5, 0.05, -0.02, 0.1) # fixed effects coefficients non-zero part
shape <- 2 # shape/size parameter of the negative binomial distribution
gammas <- c(-2.5, 0.5, 0.01) # fixed effects coefficients zero part
D11 <- 0.7 # variance of random intercepts non-zero part
D22 <- 0.5 # variance of random slopes non-zero part
D33 <- 0.5 # variance of random intercepts zero part

# we simulate random effects
b <- cbind(rnorm(n, sd = sqrt(D11)), rnorm(n, sd = sqrt(D22)), rnorm(n, sd = sqrt(D33)))
# linear predictor non-zero part
eta_y <- as.vector(X %*% betas + rowSums(Z * b[DF$id, 1:2, drop = FALSE]))
# linear predictor zero part
eta_zi <- as.vector(X_zi %*% gammas + rowSums(Z_zi * b[DF$id, 3, drop = FALSE]))
# we simulate negative binomial longitudinal data
DF$y <- rnbinom(n * K, size = shape, mu = exp(eta_y))
# we set the extra zeros
DF$y[as.logical(rbinom(n * K, size = 1, prob = plogis(eta_zi)))] <- 0
# set MCAR missing data
DF$y[sample(nrow(DF), nrow(DF) * 0.18)] <- NA
DF$y[DF$y > 100] <- NA
DF <- DF[!is.na(DF$y), ]
ls <- ls()
rm(list = c(ls[!ls %in% "DF"], "ls"))
```

# Introduction

## Over-Dispersion

## Zero-Inflation

# An Example
Function `mixed_model()` of the **GLMMadaptive** package can also be used to fit zero-inflated and two-part mixed effects models. For both types of models, a suitable `family` object needs to be specified as outlined in `vignette("Custom_Models", package = "GLMMadaptive")`, and also arguments `zi_fixed` and `zi_random` of `mixed_model()` come into play. In these 
arguments, the user can specify the fixed and random effects `formulas` of the logistic 
regression for the zero-part of the distribution of the outcome. We should note that the 
user has the option to leave `zi_random` set to `NULL`, in which case for the zero-part
we have a logistic regression with only fixed effects and no random effects.


## Descriptives
We will start with descriptive plots to see how the data look like. First, the average longitudinal evolutions
```{r dscr_marg_evol, fig.align = "center", fig.width = 8.5, fig.height = 7.5}
library("lattice")
xyplot(y ~ time | sex, data = DF, 
       panel = function (...) {
           panel.xyplot(...)
           panel.loess(..., lwd = 2, col = "red")
       })
```

## Modeling
Our primary aim here is to illustrate the extensions of the Poisson mixed model described above, to account for over-dispersion and extra zeros. Hence, we will not be focusing on model building and in particular in selecting the appropriate fixed- and random-effects structure.

We start by fitting a Poisson mixed model to the dataset using the `mixed_model()` function from the **GLMMadaptive** package:
```{r Poisson}
fm1 <- mixed_model(y ~ sex * time + age, random = ~ time | id, data = DF,
                   family = poisson())
```

## Goodness-of-Fit

## Effect Plots




