---
title: "Count Data Mixed Models - Extensions"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library("GLMMadaptive")
library("splines")
library("lattice")
library("DHARMa")
data("aids", package = "JM")
aids$CD4count <- aids$CD4 * aids$CD4
```


# Introduction

## Over-Dispersion

## Zero-Inflation

# An Example
Function `mixed_model()` of the **GLMMadaptive** package can also be used to fit zero-inflated and two-part mixed effects models. For both types of models, a suitable `family` object needs to be specified as outlined in `vignette("Custom_Models", package = "GLMMadaptive")`, and also arguments `zi_fixed` and `zi_random` of `mixed_model()` come into play. In these arguments, the user can specify the fixed and random effects `formulas` of the logistic regression for the zero-part of the distribution of the outcome. We should note that the 
user has the option to leave `zi_random` set to `NULL`, in which case for the zero-part
we have a logistic regression with only fixed effects and no random effects.


## Descriptives
We will start with descriptive plots to see how the data look like. First, the average longitudinal evolutions
```{r dscr_marg_evol, fig.align = "center", fig.width = 8.5, fig.height = 7.5}
xyplot(CD4count ~ obstime | drug, data = aids, 
       panel = function (...) {
           panel.xyplot(...)
           panel.loess(..., lwd = 2, col = "red")
       })
```

## Modeling
Our primary aim here is to illustrate the extensions of the Poisson mixed model described above, to account for over-dispersion and extra zeros. Hence, we will not be focusing on model building and in particular in selecting the appropriate fixed- and random-effects structure.

We start by fitting a Poisson mixed model to the dataset using the `mixed_model()` function from the **GLMMadaptive** package:
```{r Poisson}
fm1 <- mixed_model(CD4count ~ drug * obstime, random = ~ obstime | patient, 
                   data = aids, family = poisson())
```

```{r Netative_Binomial}
fm2 <- mixed_model(CD4count ~ drug * obstime, random = ~ obstime | patient, 
                   data = aids, family = negative.binomial())
```

```{r LRT_Pois_NB}
anova(fm1, fm2)
```

```{r ZI_Poisson}
gm1 <- mixed_model(CD4count ~ drug * obstime, random = ~ obstime | patient, 
                   data = aids, family = zi.poisson(),
                   zi_fixed = ~ drug + obstime)
```

```{r ZI_Netative_Binomial}
gm2 <- mixed_model(CD4count ~ drug * obstime, random = ~ obstime | patient, 
                   data = aids, family = zi.negative.binomial(),
                   zi_fixed = ~ drug + obstime)
```

```{r LRT_Pois_ziPois}
anova(fm1, gm1)
```

```{r LRT_NB_ziNB}
anova(fm2, gm2)
```

```{r ZI_Poisson2}
km1 <- mixed_model(CD4count ~ drug * obstime, random = ~ obstime | patient, 
                   data = aids, family = zi.poisson(),
                   zi_fixed = ~ drug + obstime, zi_random = ~ 1 | patient)
```

```{r ZI_Netative_Binomial2}
km2 <- mixed_model(CD4count ~ drug * obstime, random = ~ obstime | patient, 
                   data = aids, family = zi.negative.binomial(),
                   zi_fixed = ~ drug + obstime, zi_random = ~ 1 | patient)
```

```{r LRT_Pois_ziPois2}
anova(gm1, km1)
```

```{r LRT_NB_ziNB2}
anova(gm2, km2)
```

## Goodness-of-Fit

```{r resids_FUN}
resids_plot <- function (object, y, nsim = 1000,
                         type = c("subject_specific", "mean_subject"),
                         integerResponse = NULL) {
    if (!inherits(object, "MixMod"))
        stop("this function works for 'MixMod' objects.\n")
    type <- match.arg(type)
    if (is.null(integerResponse)) {
        integer_families <- c("binomial", "poisson", "negative binomial",
                              "zero-inflated poisson", "zero-inflated negative binomial", 
                              "hurdle poisson", "hurdle negative binomial")
        numeric_families <- c("hurdle log-normal", "beta", "hurdle beta")
        if (object$family$family %in% integer_families) {
            integerResponse <- TRUE
        } else if (object$family$family %in% numeric_families) {
            integerResponse <- FALSE
        } else {
            stop("non build-in family object; you need to specify the 'integerResponse',\n",
                 "\targument indicating whether the outcome variable is integer or not.\n")
        }
    }
    sims <- simulate(object, nsim = nsim, type = type)
    fits <- fitted(object, type = type)
    dharmaRes <- DHARMa::createDHARMa(simulatedResponse = sims, observedResponse = y, 
                              fittedPredictedResponse = fits, 
                              integerResponse = integerResponse)
    DHARMa:::plot.DHARMa(dharmaRes, quantreg = FALSE)
}
```

```{r resid_plot_Pois, fig.align = "center", fig.width = 8.5, fig.height = 7.5}
resids_plot(fm1, aids$CD4count)
```

```{r resid_plot_NB, fig.align = "center", fig.width = 8.5, fig.height = 7.5}
resids_plot(fm2, aids$CD4count)
```

```{r resid_plot_ziPois2, fig.align = "center", fig.width = 8.5, fig.height = 7.5}
resids_plot(gm1, aids$CD4count)
```

```{r resid_plot_ziNB2, fig.align = "center", fig.width = 8.5, fig.height = 7.5}
resids_plot(gm2, aids$CD4count)
```

```{r resid_plot_ziPois3, fig.align = "center", fig.width = 8.5, fig.height = 7.5}
resids_plot(km1, aids$CD4count)
```

```{r resid_plot_ziNB3, fig.align = "center", fig.width = 8.5, fig.height = 7.5}
resids_plot(km2, aids$CD4count)
```

## Effect Plots

```{r plot_data}
nDF <- with(aids, expand.grid(drug = levels(drug),
                              obstime = seq(0, 18, length.out = 11)))

plot_data <- effectPlotData(gm2, nDF)
```

```{r effect_plot, fig.align = "center", fig.width = 8.5, fig.height = 7.5}
my.panel.bands <- function(x, y, upper, lower, fill, col, subscripts, ..., font, 
                           fontface) {
    upper <- upper[subscripts]
    lower <- lower[subscripts]
    panel.polygon(c(x, rev(x)), c(upper, rev(lower)), col = fill, border = FALSE, ...)
}

xyplot(exp(pred) ~ obstime, group = drug, data = plot_data, 
       upper = exp(plot_data$upp), low = exp(plot_data$low), 
       type = "l", col = c("blue", "red"), 
       fill = c("#0000FF80", "#FF000080"),
       panel = function (x, y, ...) {
           panel.superpose(x, y, panel.groups = my.panel.bands, ...)
           panel.xyplot(x, y, lwd = 2,  ...)
}, xlab = "Follow-up time", ylab = "Expected Counts", ylim = c(0, 50))
```


